An Object-Logic Explanation for the Inconsistency in G\"odel's Ontological Theory (Extended Abstract)

### 2016-07-11 Papers/2016/KI/main.tex
Fix typos and make minor grammatical improvements

Scott's version of G\"odel's argument was automatically reconstructed by the higher-order theorem prover \textsc{Leo-II} \cite{C40} and its correctness was verified step-by-step in the \textsc{Coq} proof assistant \cite{CSR}. To bridge the gap between higher-order logics (HOL; cf.~\cite{andrewsSEP} and the references
therein), as used by these systems, and higher-order \emph{modal} logics (HOML; cf.~\cite{homl} and the
references therein), on which the ontological argument argument, the logic embedding approach \cite{J23,C40} was used.

However, G\"odel's original axioms, as used in his
manuscript \cite{GoedelNotes}, are inconsistent. This fact has
remained unnoticed to philosophers until 2013, 
when \textsc{Leo-II} found a surprising refutation of the axioms.
In \cite{C55} we extracted from Leo-II's machine-oriented refutation an informal and human-oriented intuitive explanation for the inconsistency, and we reconstructed and
verified it in the \textsc{Isabelle} proof assistant. But that explanation relied on reasoning at the meta-logic (HOL) level, which was only possible because of the embedding. Here we complement that work with a purely object-logic (HOML) explanation, and we compare and formalize both explanations in the \textsc{Coq} proof assistant.

@@

\section{An Essential Difference in the Definitions of Essence.}
\label{sec:history}
G\"odel's manuscript can be considered a translation of Leibniz's
ideas on the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in \textsc{Isabelle}, is shown in
Fig.~\ref{Scott_Goedel}. 
The main difference to G\"odel's version is an
extra conjunct in the definition of \emph{essence} (\emph{ess}). For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required.

G\"odel's omission has been
considered inessential and merely an oversight by many. 
For more than four decades, its serious consequences remained unnoticed, 
despite numerous analyses and criticisms of the
argument.
However, as explained here, the extra conjunct is in
fact crucial. Without it, G\"odel's original axioms are
inconsistent. With it, Scott's axioms are consistent (cf.~Fig.~\ref{Scott_Goedel},
where the model finder Nitpick \cite{Nitpick} confirms consistency). In
personal communication, Dana Scott confirmed that he was unaware
that G\"odel's original axioms were inconsistent.

@@

In our experiments in this branch of metaphysics we
utilise an embedding of HOMLs, such as \textbf{K}, \textbf{KB} and
\textbf{S5} with various domain conditions (possibilist and actualist
quantification), in HOL. More precisely, formulas in HOML are \emph{lifted}, i.e., converted
into predicates over worlds, which are themselves explicitly
represented as terms. The logical constants of HOML are translated to
HOL terms in such a way that, for instance,
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_0$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation,
%\cite{DBLP:journals/logcom/Ohlbach91}, 
which is here intra-logically realized --- and extended for
quantifiers --- in HOL by stating a set of equations defining the
logical constants. The resulting object logic is
the HOML \textbf{K} with rigid terms and constant domains (possibilist
quantifiers). Other logics (e.g. \textbf{KB}, \textbf{S5}) are
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by
using an existence predicate to guard the quantifiers.

@@

Although \textsc{Leo-II}'s resolution refutation is not easy to read
for humans, it did contain relevant hints to the importance of the
empty property\footnote{
  Note that the terms for the empty property ($\lambda x. \bot$) and for the property of self-difference ($\lambda x.  x\not=x$) have identical denotations in the logic setting
  with functional and Boolean extensionality assumed
  here. 
  For the proof to go through it is
  irrelevant which property is used.
} $\lambda x. \bot$ (also denoted $\emptyset$, as in HOL it is customary to think of unary predicates as sets)\footnote{An additional lambda abstraction occurs in the empty property in \textsc{Leo-II}'s proof (and in the reconstruction in \textsc{Isabelle}) because the embedding approach lifts the boolean type $o$ to $\iota \imp o$.}. Based on this hint, we conceived the following informal explanation for the inconsistency of G\"odel's axioms (reproduced without change from \cite{C55}):
%

@@

If we were to convert the informal proof above to a formal proof, the semantic reasoning in step 8 would require a leap to the meta-logic (HOL), in order to expand the definitions of the modal operators and reason directly about possible worlds. The following alternative proof avoids this leap and remains purely within the object logic (HOML \K):

\begin{enumerate}
\item[8\textsuperscript{*}.] We must derive $\bot$ from $\pos \nec \bot$. In order to derive $\bot$, it suffices to show that there exists a derivable proposition such that its negation is also derivable. We choose $\pos \pos \bot$ as a candidate proposition, and hence we must show that:
\begin{enumerate}
\item $\neg \pos \pos \bot$ \emph{is derivable:} this proposition is equi-valid to $\nec \nec \top$, which is trivially derivable from $\top$ by two applications of the necessitation inference rule.

@@

\noindent
Due to a deliberate and disciplined use of only 
the simplest (and non-automatic) \textsc{Coq} tactics, 
there is a straightforward correspondence between the tactics used 
in the scripts above and the inference rules of the modal natural 
deduction calculus \cite{CSR}. Therefore, the lengths of the proof 
scripts (in number of tactic applications) can serve as estimations 
for the lengths of the corresponding natural deduction proof. 
It is noticeable that the meta-logic proof is significantly shorter 
than the pure object-logic proof. An in-depth analysis 
reveals that the reasoning about the possible worlds semantics in 
the meta-logic proof acts as a short-cut: when it becomes impossible (in step 8) to build the third world $w'$ (because $\bot$ would have to hold in it, and thus $w'$ would be contradictory), a contradiction at the HOL level can be immediately derived, completing the proof. In the object-level proof, on the other hand, such a contradiction had to be found in the arbitrary initial world $w$. This not only requires additional tedious logical inferences (cf. the proofs of the lemmas \texttt{mimplies\_to\_mnot} and \texttt{dia\_not\_not\_box}), but also a non-trivial guessing of the instantiating contradictory proposition $\pos \pos \bot$, whose purpose is precisely to carry over the contradiction from $w'$ back to $w$.

>>>>

Scott's version of G\"odel's argument was automatically reconstructed by the higher-order theorem prover \textsc{Leo-II} \cite{C40} and its correctness was verified step-by-step in the \textsc{Coq} proof assistant \cite{CSR}. To bridge the gap between higher-order logics (HOL; cf.~\cite{andrewsSEP} and the references
therein), as used by these systems, and higher-order \emph{modal} logics (HOML; cf.~\cite{homl} and the
references therein), on which the ontological argument is based, the logic embedding approach \cite{J23,C40} was used.

However, G\"odel's original axioms, as used in his
manuscript \cite{GoedelNotes}, are inconsistent. This fact has
remained unnoticed to philosophers until 2013, 
when \textsc{Leo-II} found a surprising refutation of the axioms.
In \cite{C55} we extracted from Leo-II's machine-oriented refutation an informal and human-oriented intuitive explanation for the inconsistency, and we reconstructed and
verified it in the \textsc{Isabelle} proof assistant. But that explanation relied on reasoning at the \emph{meta-logic} (HOL) level, which was only possible because of the embedding. Here we complement that work with a purely \emph{object-logic} (HOML) explanation, and we compare and formalize both explanations in the \textsc{Coq} proof assistant.

@@

\section{An Essential Difference in the Definitions of Essence}
\label{sec:history}

G\"odel's manuscript can be considered a translation of Leibniz's
ideas on the argument into modern modal logic. G\"odel
discussed his manuscript with Scott, who shared a slightly different
version with a larger public. Scott's version of the axioms and
definitions, formalized in \textsc{Isabelle}, is shown in
Fig.~\ref{Scott_Goedel}. 
The main difference to G\"odel's version is an
extra conjunct in the definition of \emph{essence} (\emph{ess}). For Scott,
an essential property of an individual must be possessed by
him/her. For G\"odel, this is not required.

G\"odel's omission has been
considered inessential and merely an oversight by many. 
For more than four decades, its serious consequences remained unnoticed, 
despite numerous analyses and criticisms of the
argument.
However, as explained here, the extra conjunct is in
fact crucial. Without it, G\"odel's original axioms are
inconsistent. With it, Scott's axioms are consistent (cf.~Fig.~\ref{Scott_Goedel},
where the model finder \textsc{Nitpick} \cite{Nitpick} confirms consistency). In
personal communication, Dana Scott confirmed that he was unaware
that G\"odel's original axioms were inconsistent.

@@

In our experiments in this branch of metaphysics we
utilize an embedding of HOMLs, such as \textbf{K}, \textbf{KB} and
\textbf{S5} with various domain conditions (possibilist and actualist
quantification), in HOL. More precisely, formulas in HOML are \emph{lifted}, i.e., converted
into predicates over worlds, which are themselves explicitly
represented as terms. The logical constants of HOML are translated to
HOL terms in such a way that, for instance,
%$\neg \varphi$, $\varphi\vee\psi$
$\Box \varphi$ and $\Diamond \varphi$ (relative to a current world
$w_0$) are mapped, respectively, to the HOL formulas
$\forall w. (r w_0 w) \imp (\varphi w)$ and
$\exists w. (r w_0 w) \wedge (\varphi w)$. This form of embedding is
precisely the well-known standard translation,
%\cite{DBLP:journals/logcom/Ohlbach91}, 
which is here intra-logically realized --- and extended for
quantifiers --- in HOL by stating a set of equations defining the
logical constants. The resulting logic is
the HOML \textbf{K} with rigid terms and constant domains (possibilist
quantifiers). Other logics (e.g. \textbf{KB}, \textbf{S5}) are
embedded by adding axioms that restrict the accessibility relation
$r$. Varying domains and actualist quantifiers can be simulated by
using an existence predicate to guard the quantifiers.

@@

Although \textsc{Leo-II}'s resolution refutation is not easy to read
for humans, it did contain relevant hints to the importance of the
empty property\footnote{
  Note that the terms for the empty property ($\lambda x. \bot$) and for the property of self-difference ($\lambda x.  x\not=x$) have identical denotations in the logic setting
  with functional and Boolean extensionality assumed
  here. 
  For the proof to go through, it is
  irrelevant which property is used.
} $\lambda x. \bot$ (also denoted $\emptyset$, as in HOL it is customary to think of unary predicates as sets)\footnote{An additional lambda abstraction occurs in the empty property in \textsc{Leo-II}'s proof (and in the reconstruction in \textsc{Isabelle}) because the embedding approach lifts the boolean type $o$ to $\iota \imp o$.}. Based on this hint, we conceived the following informal explanation for the inconsistency of G\"odel's axioms (reproduced without change from \cite{C55}):
%

@@

If we were to convert the informal proof above to a formal proof, the semantic reasoning in step 8 would require a leap to the meta-logic (HOL), in order to expand the definitions of modal operators and reason directly about possible worlds. The alternative proof below avoids this leap and remains purely within the object logic (HOML \K):

\begin{enumerate}
\item[8\textsuperscript{*}.] We must derive $\bot$ from $\pos \nec \bot$. In order to derive $\bot$, it suffices to show that there exists a derivable proposition such that its negation is also derivable. We choose $\pos \pos \bot$ as the candidate proposition, and hence we must show that:
\begin{enumerate}
\item $\neg \pos \pos \bot$ \emph{is derivable:} this proposition is equi-valid to $\nec \nec \top$, which is trivially derivable from $\top$ by two applications of the necessitation inference rule.

@@

\noindent
Due to a deliberate and disciplined use of only 
the simplest (and non-automatic) \textsc{Coq} tactics, 
there is a straightforward correspondence between the tactics used 
in the scripts above and the inference rules of the modal natural 
deduction calculus \cite{CSR}. Therefore, the lengths of the proof 
scripts (in number of tactic applications) can serve as estimations 
for the lengths of the corresponding natural deduction proof. 
It is noticeable that the meta-logic proof is significantly shorter 
than the pure object-logic proof. An in-depth analysis 
reveals that the reasoning about the possible worlds semantics in 
the meta-logic proof acts as a short-cut: when it becomes impossible (in step 8) to build the third world $w'$ (because $\bot$ would have to hold in it, and thus $w'$ would be contradictory), a contradiction at the HOL level can be immediately derived, completing the proof. In the object-level proof, on the other hand, such a contradiction has to be found in the arbitrary initial world $w$. This requires not only additional tedious logical inferences (cf. the proofs of the lemmas \texttt{mimplies\_to\_mnot} and \texttt{dia\_not\_not\_box}), but also a non-trivial guessing of the contradictory proposition $\pos \pos \bot$, whose purpose is precisely to carry over the contradiction from $w'$ back to $w$.
